### 2024-10-08 teaching/math240f24/draft.tex
Bezout and FTA

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
\def\divides{\mid}

@@

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}

@@

\medskip\boldlabel\datestamp{05}{ix} Russell's paradox. Earlier, we said that the sets we are working with
need to be a subset of a universe $U$, which has already been proven to be a set. We gave lots of
ways to make sets out of new sets, such as the union and intersection operations, etc. Starting with
the assumption that the empty set $\emptyset$ is a set,
it is possible to define the set of natural numbers as follows.
We can define $0 = \emptyset$, $1 = \{0\}$, and $2 = \{0,1\}$, and so on.
Now we take the set of all of these, and call this $\NN$.
(We can also define addition and multiplication on these set-theoretic ``numbers''
so that they behave like addition and multiplication do on $\NN$.) From here we can do more funky stuff
to define $\ZZ$, $\QQ$, and $\RR$, and prove that these are all sets (you can see this in a higher-level
course on set theory, if you're interested). So there isn't much of a problem
with all the sets we have played with so far; they are all subsets of things that are already known to be sets.

@@

\proof By part (i) the previous proposition, every $x\in A$ belongs to the equivalence class $[x]$.
Then, by part (ii) of the previous proposition, we know that for any equivalence classes
$[x]$ and $[y]$ such that $[x]\ne [y]$, we must have $x\not\sim y$, and by part (iii) of the previous
proposition, we deduce that $[x]\cap [y]=\emptyset$. This shows that $A/\sim$ satisfies the second
part of the definition of partition. Lastly, we note that $\emptyset\notin A/\sim$, since
every element of $A/\sim$ is equal to $[x]$ for some $x\in A$, and must thus contain at least
the element $x$.\slug

Let us now revisit the examples of equivalence relations from last class, and see what partitions
they give rise to. In the example $F$ of propositional formulas with variables $p$, $q$, and $r$,
under the equivalence relation $\equiv$, the set of equivalence classes is the set of all possible
truth tables on three variables. Each such truth table has $8$ rows, so there are $2^8$ equivalence
classes. In other words, $|F/\equiv| = 2^8$.

How about the relation $R$ defined on $\RR^2$ where
$$(x,y)\sim (w,z) \qquad\hbox{if and only if}\qquad \sqrt{x^2+y^2} = \sqrt{w^2 + z^2}?$$
Well, each $\bigl[(x,y)\bigr]$ is the circle of radius $\sqrt{x^2 + y^2}$ centred
around the origin, so $\RR^2/\sim$ is the set of all circles in the plane centred at $(0,0)$.

@@

\proof Let
$$S = \{ s'a+t'b : s',t'\in \ZZ,\ ax+by > 0\}.$$
This set is nonempty, since if $a$ is negative then $(-1)a+0b\in S$ and if $a$ is positive,
then $1a+0b\in S$. Since $S$ is a nonempty set of positive integers, it has
a least element, by the well-ordering principle. Call this integer $d = sa+tb$ (for some
specific choices of $s,t\in \ZZ$);
the claim is that
$d = \gcd(a,b)$.

@@

We now summarise this section on B\'ezout's identity with a little scenario.
Imagine a robot that lives on the line of integers. It starts at the point $0$
and can hop in steps of $a$ or $b$ (in either direction). Theorem~{\thmbezout} tells us
that the point $d = \gcd(a,b)$ is reachable by the robot. The next proposition characterises
the set of {\it all} points that the robot can get to.

@@

\proclaim Theorem \advthm. An integer $p$ with $p\ge 2$ is prime if and only if
for all $a,b\in \NN$, $p\divides ab$ implies that $p\divides a$ or $p\divides b$.

@@

\noindent\datestamp{13}{} {\it Proof.}\enspace The proof that such a decomposition of $n$ exists is
by (strong) induction. The base case is $n=2$. This is already a prime factorisation, since $2$ is prime.

@@

Next, suppose that the uniqueness
statement holds for $s$ (i.e., for any integer that decomposes
into a product of $s$ primes, the decomposition is unique) and we want to show that it holds
for $s+1$. We assume that
$$n = p_1p_2\cdots p_{s+1}$$
and
$$n = q_1q_2\cdots q_t,$$
where $p_1\le p_2\le \cdots \le p_{s+1}$ and $q_1\le q_2\le\cdots\le q_t$. We
have $p_{s+1}\divides n = q_1q_2\cdots q_t$, so by the preceding corollary, there
exists $1\le i\le t$ such that $p_{s+1}\divides q_i$, and since $q_i$ is prime, this means
that $p_{s+1} = q_i\le q_t$. Similarly, $q_t$ divides $n = q_1\ldots p_{s+1}$, so it
divides $p_j$ for some $1\le j\le s+1$. This means that $q_t = p_j \le p_{s+1}$. Hence
$q_t = p_{s+1}$. But
$$p_1p_2 \cdots p_s p_{s+1} = q_1q_2 \cdots q_t,$$
so we may divide out by $p_{s+1} = q_t$ on both sides to get
$${n\over p_{s+1}} = p_1p_2\cdots p_s = q_1 q_2 \cdots q_{t-1}.$$
By the induction hypothesis, the decomposition of $n/p_{s+1}$ is unique, so $t-1 = s$
and $p_i = q_i$ for all $1\le i\le t-1$. This proves that $s+1 = t$ and
$p_i = q_i$ for all $1\le i\le s+1$.\slug

@@

>>>>

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
\def\divides{\mid}
\def\percent{\mathrel{\%}}

@@

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}

@@

\medskip\boldlabel\datestamp{05}{ix} Russell's paradox. Earlier, we said that the sets we are working with
need to be a subset of a universe $U$, which has already been proved to be a set. We gave lots of
ways to make sets out of new sets, such as the union and intersection operations, etc. Starting with
the assumption that the empty set $\emptyset$ is a set,
it is possible to define the set of natural numbers as follows.
We can define $0 = \emptyset$, $1 = \{0\}$, and $2 = \{0,1\}$, and so on.
Now we take the set of all of these, and call this $\NN$.
(We can also define addition and multiplication on these set-theoretic ``numbers''
so that they behave like addition and multiplication do on $\NN$.) From here we can do more funky stuff
to define $\ZZ$, $\QQ$, and $\RR$, and prove that these are all sets (you can see this in a higher-level
course on set theory, if you're interested). So there isn't much of a problem
with all the sets we have played with so far; they are all subsets of things that are already known to be sets.

@@

\proof By part (i) the previous proposition, every $x\in A$ belongs to the equivalence class $[x]$.
Then, by part (ii) of the previous proposition, we know that for any equivalence classes
$[x]$ and $[y]$ such that $[x]\ne [y]$, we must have $x\not\sim y$, and by part (iii) of the previous
proposition, we deduce that $[x]\cap [y]=\emptyset$. This shows that $A/{\sim}$ satisfies the second
part of the definition of partition. Lastly, we note that $\emptyset\notin A/{\sim}$, since
every element of $A/{\sim}$ is equal to $[x]$ for some $x\in A$, and must thus contain at least
the element $x$.\slug

Let us now revisit the examples of equivalence relations from last class, and see what partitions
they give rise to. In the example $F$ of propositional formulas with variables $p$, $q$, and $r$,
under the equivalence relation $\equiv$, the set of equivalence classes is the set of all possible
truth tables on three variables. Each such truth table has $8$ rows, so there are $2^8$ equivalence
classes. In other words, $|F/{\equiv}| = 2^8$.

How about the relation $R$ defined on $\RR^2$ where
$$(x,y)\sim (w,z) \qquad\hbox{if and only if}\qquad \sqrt{x^2+y^2} = \sqrt{w^2 + z^2}?$$
Well, each $\bigl[(x,y)\bigr]$ is the circle of radius $\sqrt{x^2 + y^2}$ centred
around the origin, so $\RR^2/{\sim}$ is the set of all circles in the plane centred at $(0,0)$.

@@

\noindent\datestamp{08}{x} {\it Proof.}\enspace
Let
$$S = \{ s'a+t'b : s',t'\in \ZZ,\ ax+by > 0\}.$$
This set is nonempty, since if $a$ is negative then $(-1)a+0b\in S$ and if $a$ is positive,
then $1a+0b\in S$. Since $S$ is a nonempty set of positive integers, it has
a least element, by the well-ordering principle. Call this integer $d = sa+tb$ (for some
specific choices of $s,t\in \ZZ$);
the claim is that
$d = \gcd(a,b)$.

@@

We now summarise this section on B\'ezout's identity with a little scenario.
Imagine a frog that lives on a doubly-infinite line of lilypads,
indexed by the integers. It starts at the point $0$
and can hop in steps of $a$ or $b$ (in either direction). Theorem~{\thmbezout} tells us
that the lilypad $d = \gcd(a,b)$ is reachable by the frog. The next proposition characterises
the set of {\it all} lilypads that the frog can get to.

@@

\edef\thmpdivides{\the\thmcount}
\proclaim Theorem \advthm. An integer $p$ with $p\ge 2$ is prime if and only if
for all $a,b\in \NN$, $p\divides ab$ implies that $p\divides a$ or $p\divides b$.

@@

\proof The proof that such a decomposition of $n$ exists is
by (strong) induction. The base case is $n=2$. This is already a prime factorisation, since $2$ is prime.

@@

Next, suppose that the uniqueness
statement holds for $s$ (i.e., for any integer that decomposes
into a product of $s$ primes, the decomposition is unique) and we want to show that it holds
for $s+1$. We assume that
$$n = p_1p_2\cdots p_{s+1}$$
and
$$n = q_1q_2\cdots q_t,$$
where $p_1\le p_2\le \cdots \le p_{s+1}$ and $q_1\le q_2\le\cdots\le q_t$. We
have $p_{s+1}\divides n = q_1q_2\cdots q_t$, so by the preceding corollary, there
exists $1\le i\le t$ such that $p_{s+1}\divides q_i$, and since $q_i$ is prime, this means
that $p_{s+1} = q_i\le q_t$. Similarly, $q_t$ divides $n = q_1\ldots p_{s+1}$, so it
divides $p_j$ for some $1\le j\le s+1$. This means that $q_t = p_j \le p_{s+1}$. Hence
$q_t = p_{s+1}$. But
$$p_1p_2 \cdots p_s p_{s+1} = q_1q_2 \cdots q_t,$$
so we may divide out by $p_{s+1} = q_t$ on both sides to get
$${n\over p_{s+1}} = p_1p_2\cdots p_s = q_1 q_2 \cdots q_{t-1}.$$
By the induction hypothesis, the decomposition of $n/p_{s+1}$ is unique, so $t-1 = s$
and $p_i = q_i$ for all $1\le i\le t-1$. This proves that $s+1 = t$ and
$p_i = q_i$ for all $1\le i\le s+1$.\slug
\bye

@@

Next, we show that there are infinitely many primes, a theorem first proved by Euclid.

\proclaim Theorem \advthm. There are infinitely many prime numbers.

\proof Suppose, towards a contradiction, that there are finitely many prime numbers, call
them $p_1, p_2, \ldots, p_m$. Let $n = p_1p_2\cdots p_m$, and consider the integer $n+1$. Either
it is prime or it is not. If $n+1$ is prime, then we already have a contradiction, since $n+1 > p_i$
for all $1\le i\le m$, and we assumed that the $p_1,\ldots,p_m$ were all the primes. If $n+1$
is not prime, then it is divisible by some prime in our list, call it $p_i$. Hence we can write
$n = kp_i$ for some integer $k$, and we have
$$ k p_i = n+1 = p_1 p_2\cdots p_m + 1.$$
Rearranging this a bit, we have
$$ 1 = kp_i - p_1 p_2\cdots p_m = p_i(k - p_1 p_2 \cdots p_{i-1} p_{i+1} \cdots p_m);$$
in other words, $p_i$ divides $1$. This is a contradiction as $1$ is not divisible by any integer
greater than $1$.\slug

Though the set $P$ of prime numbers is infinite, it does sort of get ``sparser'' as one heads off
towards infinity. This is quantified by the following theorem, proved independently
in~1896 by Jacques Hadamard and Charles Jean de la Vall\'ee Poussin.

\parenproclaim Theorem {\advthm} (Prime number theorem). For $x\in \RR$, let
$$\pi(x) = \bigl| \{ p \le x : p\ \hbox{prime}\} \bigr|.$$
Then $pi(x) \sim x/\ln x$, in the sense that
$$\lim_{x\to\infty} {\pi(x)\over x/\ln x} = 1.$$

The proof is long and arduous, requiring a lot of background in complex analysis.
It is often taught in a first-year graduate class in analytic number theory.
Here is an easy corollary of the prime number theorem.

\proclaim Corollary \advthm. Let $n$ be a positive integer and let $m$ be chosen
uniformly at random from the set $\{1,2,\ldots, n\}$. Then
$$(\ln n)\pr\{m\ \hbox{prime}\}\to 1$$
as $n\to\infty$. In other words, the probability that $m$ is prime is asymptotically
$1/\ln n$.

\proof Since $m$ is chosen uniformly at random from $\{1,\ldots,n\}$,
the probability $\pr\{ m\ \hbox{prime}\}$ equals $\pi(n)/ n$. So
$$\lim_{n\to\infty} (\ln n) \pr\{m\ \hbox{prime}\} = \lim_{n\to\infty} {\pi(n) \ln n\over n} 
= 1,$$
by the prime number theorem.\slug

As an example, if we choose a $30$-digit number at random (so $n=10^{30} - 1$), the
probability that this number is prime is roughly $1/\ln(10^{30}) = 1/(30\ln 10) = 0.0145$
or $1.45\%$.

\advsect Modular Arithmetic

Fix $n\ge 1$, and let $a,b\in\ZZ$. We say {$a$ is congruent to $b$ modulo $n$}
if $n\divides a-b$, i.e., if $a-b = kn$ for some $k\in \ZZ$. Write $a \equiv b\pmod n$
or $a \equiv_n b$.

Take for example $n=12$. We have, e.g., $4\equiv 16\pmod 14$, since $4-16 = -12 = (-1)12$,
but, e.g., $7\not\equiv 17$, since $7-17 = -10$, and $12$ does not divide $-10$. This
situation should be a familiar one, since we are used to working with numbers modulo $12$ when
telling the time.

For any fixed $n$, the set of all $(a,b)\in \ZZ^2$ with $a\equiv_n b$ is a relation on $\ZZ$.
In fact, we have the following proposition.

\proclaim Proposition \advthm.
For all fixed $n$, the relation $\equiv_n$ is an equivalence relation on the set $\ZZ$.

\proof We must show that $\equiv_n$ is reflexive, symmetric, and transitive.

Let $a\in \ZZ$. We have $a-a = 0 = 0\cdot n$, so $a\equiv a\pmod n$, proving reflexivity.

Let $a,b\in \ZZ$ and suppose that $a\equiv_n b$, so there exists $k$ such that $a-b=kn$.
Then $b-a = (-k)n$, so $b\equiv_n a$. This proves symmetry.

Lastly, let $a,b,c\in \ZZ$ be such that $a\equiv_n b$ and $b\equiv_n c$. So there
exist integers $k,l\in \ZZ$ such that $a-b = kn$ and $b-c = ln$. Then
$$a - c = (a-b) + (b-c) = kn +ln = (k+l)n,$$
which shows that $a\equiv_n c$.\slug

Since $\equiv_n$ is an equivalence relation, it partitions $\ZZ$ into equivalence classes.
We shall denote by $[a]_n$ the equivalence class of $a$ modulo $n$. This is the set
$$[a]_n = \{b\in \ZZ : a-b = kn\ \hbox{for some}\ k\in \ZZ\}
= \{a + ln : l\in \ZZ\}.$$
For instance, when $n = 3$, the set
$$[0]_3 = \{\ldots, -6, -3, 0, 3, 6, \ldots\}$$
is just the set of all multiples of $3$, and we also have
$$[1]_3 = \{\ldots, -5, -2, 1, 4, 7,\ldots\}$$
and
$$[2]_3 = \{\ldots, -4, -1, 2, 5, 8,\ldots\}.$$
These are all of the equivalence classes, since, for instance,
$$[4]_3 = \{\ldots, -2, 1, 4, 7, 10,\ldots\} = [1]_3.$$

Lastly, we shall touch upon the {\it modulo} operator, which is a feature of many programming
languages. Let $a\in \ZZ$ and $b\ge 1$. Let $a \percent b = r$ where $q$ and $r$ are the integers
given by the division algorithm. (That is, $a = qb+r$ where $0\le r\le b$.)

\proclaim Proposition \advthm. Fix an integer $n\ge 2$.
Let $a,b\in \ZZ$. Then $a\equiv b\pmod n$ if and only
if
$$a\percent n = b\percent n.$$

\noindent\datestamp{14}{} {\it Proof.}\enspace
Suppose that $a\equiv_n b$, so that there exists $k\in \NN$ such that $a-b = kn$.
Then, let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
(So $r_1 = a\percent n$ and $r_2 = b\percent n$, and we have $0\le r_1,r_2 < n$.)
Write
$$kn = a - b = (q_1-q_2) n + (r_1 - r_2),$$
which we can rearrange to
$$r_1 - r_2 = kn - (q_1 - q_2)n = (k-q_1 + q_2)n.$$
This implies that $r_1 - r_2$ divides $n$, but
since $r_1, r_2\in [0,n)$, the quantity $r_1 - r_2$ is in the range $[-n-1, n)$. Hence
the only way it can divide $n$ is for $r_1 - r_2 = 0$. We conclude that $a\percent n = b\percent n$.

For the other direction, we once again
let $a = nq_1 + r_1$ and $b = nq_2 + r_2$ from the division algorithm.
Now the assumption is that $r_1 = a\percent n = b\percent n = r_2$. So
$$a - b = a - b = (q_1-q_2) n + (r_1 - r_2) = (q_1-q_2)n.$$
This shows that $a\equiv b\pmod n$.\slug

This proposition is useful in practice. For example, suppose we want to know whether
$22\equiv 8\pmod 3$. We calculate $22 = 7\cdot 3 + 1$, so $22\percent 3 = 1$; meanwhile
$8 = 2\cdot 3 + 2$, so $8\percent 3 = 2$. By the proposition, this means that
$22\not\equiv 8\pmod 3$.

The proposition also implies that for all $a\in \ZZ$ and $n\ge 2$ one has $[a]_n = [a\percent n]$.
From the division algorithm, we know that $a\percent n$ is an element in the range $[0,n)$;
it is equal to the integer $r$ in that range such that we may write $a = qn+r$ for some
integer $q$. We may choose to denote the whole equivalence class by this element $r$. The set
$$\ZZ/n\ZZ = \ZZ/{\equiv_n} = \bigl\{ [0], [1], \ldots, [n-1] \bigr\}$$
is called the {\it ring of integers modulo $n$} or the {\it cyclic group on $n$ elements}.

\medskip\boldlabel Computations modulo {\mathbold n}.
We can do addition and multiplication modulo $n$ by first doing addition and multiplication
and then taking the remainder with respect to $n$. For example, when $n=4$ we have the
addition table
$$
\vbox{\ninepoint\tabskip7pt\offinterlineskip
\halign{
\hfil $#$ \hfil & \hfil\vrule# \hfil \enskip &
\hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
+ && 0 & 1 & 2 & 3 \cr
& height7pt &&&& \cr
\noalign{\hrule}
& height7pt &&&& \cr
0 && 0 & 1 & 2 & 3 \cr
& height7pt &&&& \cr
1 && 1 & 2 & 3 & 0 \cr
& height7pt &&&& \cr
2 && 2 & 3 & 0 & 1 \cr
& height7pt &&&& \cr
3 && 3 & 0 & 1 & 2 \cr
}}
$$
and the multiplication table
$$
\vbox{\ninepoint\tabskip7pt\offinterlineskip
\halign{
\hfil $#$ \hfil & \hfil\vrule# \hfil \enskip &
\hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil & \hfil $#$ \hfil \cr
\cdot && 0 & 1 & 2 & 3 \cr
& height7pt &&&& \cr
\noalign{\hrule}
& height7pt &&&& \cr
0 && 0 & 0 & 0 & 0 \cr
& height7pt &&&& \cr
1 && 0 & 1 & 2 & 3 \cr
& height7pt &&&& \cr
2 && 0 & 2 & 0 & 2 \cr
& height7pt &&&& \cr
3 && 0 & 3 & 2 & 1 \cr
}}.
$$
For $x,y\in \ZZ$, if $xy = 0$, then either $x=0$ or $y=0$.
On the other hand, from the assumption $xy \equiv 0\pmod n$ we cannot conclude
in general that $x\equiv 0\pmod n$ or $y\equiv 0\pmod n$. As an example, when $n=4$
and $x=y=2$, neither $x$ nor $y$ is congruent to $0$ modulo $4$, but their product is $0$ modulo $4$.
An element $a\in \ZZ$ with $a\not\equiv 0\pmod n$ is said to be a {\it zero divisor}
if there exists $b\in \ZZ$ with $b\not\equiv 0\pmod n$ such that $ab \equiv 0\pmod n$.

Remember that the numbers in the tables above are elements of $\ZZ/n\ZZ$ (in the case $n=4$),
hence they are not really integers so much as equivalence classes of integers, which are sets!
So it is not immediate that addition and multiplication are well-defined modulo some fixed integer $n$.
For instance, to say that $[3]_4 \cdot [1]_4 = [3]_4$, it is not enough to just observe that
$3\cdot 1 = 3$. We need to show that for any integers $a\in [3]_4$ and $b\in [1]_4$, we have
$a\cdot b\in [3]_4$. This is the essence of the following proposition.

\proclaim Proposition \advthm. Let $a\equiv c\pmod n$ and $b\equiv d\pmod n$. Then
\medskip
\item{i)} $a+b\equiv c+d \pmod n$;
\smallskip
\item{ii)} $ab\equiv cd\pmod n$; and
\smallskip
\item{iii)} $a^m \equiv c^m \pmod n$ for all $m\in \NN$.
\medskip

\proof By assumption, there are integers $k$ and $l$ such that $a-c = kn$ and $b-d = ln$.

To prove (i), we compute
$$(a+b)-(c+d) = a-c + b-d = kn+ln = (k+l)n,$$
so $a+b\equiv c+d\pmod n$.

For part (ii), we compute
$$\eqalign{
ab-cd &= ab-cb+cb - cd \cr
&= b(a-c) + c(b-d) \cr
&= bkn + cln \cr
&= (bk + cl) n ,\cr
}$$
so $ab\equiv cd \pmod n$.

We prove part (iii) by induction. The base case is $m=0$; we have $a^0 = 1 = c^0$,
so $a^0\equiv c^0\pmod n$. Now let $m\ge 0$ and assume that $a^m\equiv c^m\pmod n$.
Combining this with the hypothesis $a\equiv c\pmod n$, we use part (ii) to
conclude that $a\cdot a^m \equiv c\cdot c^m\pmod n$, which simplifies
to $a^{m+1}\equiv c^{m+1}\pmod n$.\slug

We can use this to perform the modulo operation on large numbers without necessarily
knowing the large numbers themselves. For example, let $m=21^{10} - 3\cdot 13 + 14$
and suppose we want to know $m\percent 6$.

First we observe that $14\equiv 2\pmod 6$
and $13\equiv 1\pmod 6$. So
$$ m = 21^{10} - 3\cdot 13 + 14 \equiv_6 21^{10} - 3\cdot 1 + 2 \equiv_6 21^{10} - 1.$$
Now we deal with the exponent by first taking the base of the exponent modulo $6$.
We have $21\equiv 3 \pmod 6$; hence
$$m \equiv_6 3^{10} -1.$$
To reduce the exponent,
we factorise $10=2\cdot 5$, yielding
$$m \equiv_6 (3^2)^5 - 1.$$
But again we can reduce the base here, since $9\equiv_6 3$, which gives us
$$m \equiv_6 3^5-1.$$
Now since the exponent $5$ is prime, we aren't able to reduce it by factorising. However,
we do have $5 = 2+2+1$, and, using once again the fact that $3^2 = 9\equiv_6 3$, we obtain
$$m \equiv_6 3^2\cdot 3^2 \cdot 3 - 1 \equiv_6 3\cdot 3\cdot 3 - 1 \equiv_6 3\cdot 3 - 1
\equiv_6 3- 1 \equiv_6 2.$$

\medskip\boldlabel Inverses modulo {\mathbold n}. Considering $2$ as a rational
number, we say that the inverse of $2$ is $1/2$, since $2\cdot (1/2) = 1$.
We write $2^{-1} = 1/2$. We shall use the same terminology in the ring of integers modulo $n$.
An element $a\in \ZZ$ is said to be {\it invertible modulo $n$} if there exists $b\in \ZZ$
such that $ab \equiv 1\pmod n$. In this case we shall say that $b$ is an {\it inverse} of $a$.
In fact, inverses are unique (in $\ZZ/n\ZZ$).

\proclaim Proposition \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then if $ab \equiv 1\pmod n$
and $ac\equiv 1\pmod n$, then $b\equiv c\pmod n$.

\proof We have
$$b \equiv b\cdot 1 \equiv b(ac) \equiv (ab)c \equiv 1\cdot c \equiv c\pmod n.\noskipslug$$

Hence we may speak of {\it the} inverse of an element $a$, which we shall denote by $a^{-1}$.
We have $a^{-1}a \equiv 1 \equiv a a^{-1}\pmod n$.

For instance, when $n=5$, the fact that
$$2\cdot 3 \equiv_5 6\equiv_5 1$$
tells us that $2$ (and $3$, for that matter) are invertible modulo $5$.
When $n=6$, is $2$ still invertible? Here $n$ is small enough that we can just try all the cases.
We have $2\cdot 0 = 0$, $2\cdot 1 = 2$, $2\cdot 2 = 4$, $2\cdot 3 = 6\equiv_6 = 0$
$2\cdot 4 = 8 \equiv_6 2$, and $2\cdot 5 = 10\equiv_6 4$. Since none of these were equivalent
to $1$ modulo $6$, we see that $2$ is not invertible modulo $6$.

It was quite a tedious process to exhaustively show that $2$ is not invertible modulo $6$. The following
theorem gives a much easier-to-use criterion for invertibility.

\proclaim Theorem \advthm. Let $a,n\in \ZZ$ with $n\ge 2$. Then
\medskip
\item{i)} $a$ is invertible modulo $n$ if and only if $\gcd(a,n) = 1$; and
\smallskip
\item{ii)} if $a$ is invertible, then there is a unique integer $b\in [0,n-1]$ such that
$ab\equiv 1\pmod n$. Namely, if
$$1 = sa + tn$$,
then we can set $b = s\percent n$.
\medskip

\proof The proof of (i) is a fairly simple corollary of B\'ezout's theorem.
f $a$ is invertible, then $ab\equiv 1\pmod n$, so there exists $k\in \ZZ$ such that
$1-ab=kn$. Rearranging this, we have $1 = ab+kn$, and by B\'ezout's theorem (specifically,
the final statement of Theorem~{\thmbezout}, this implies that $\gcd(a,n) = 1$.
On the other hand, if $\gcd(a,n) = 1$, then there exists $k\in \ZZ$ such that
$1 = ab + kn$. This means that $1-ab = kn$, so $ab\equiv 1\pmod n$.

For part (ii), suppose that $s,t\in \ZZ$ are such that $1 = sa+tn$. Then
taking this equation modulo $n$, we have
$$ 1 = sa+tn \equiv_n sa + t\cdot 0 \equiv_n sa.$$
Letting $b = s\percent n$, we have $1 = ba$, since $s$ and $b$ belong to the same
equivalence class modulo $n$, and $b$ is an integer in $[0,n-1]$ by the division algorithm.
This proves existence of such a $b$. To show uniqueness, suppose that $b$ and $c$ are both
integers in $[0,n-1]$ with $ab \equiv 1\pmod n$ and $ac\equiv 1\pmod n$. Then
$ab\equiv ac\pmod n$, and multiplying on both sides by $b$, we have $bab\equiv bac\pmod n$.
But $ba\equiv 1\pmod n$, so cancelling this out in the equation, we have $b\equiv c\pmod n$.
In other words, $n\divides b-c$. But since $b,c\in [0,n-1]$, we have $b-c\in [-n+1, n-1]$,
and the only divisor of $n$ in this range is $0$. Hence $b-c = 0$ and we conclude that $b=c$.\slug

We can use this theorem to systematically find inverses of integers modulo other integers. For instance,
suppose we want to find the inverse of $17$ modulo $20$. First we find $\gcd(20,17)$ by Euclid's
algorithm:
$$\eqalign{
20 &= 1\cdot 17 + 3 \cr
17 &= 5\cdot 3 + 2 \cr
3 &= 1\cdot 2 + 1 \cr
1 &= 1\cdot 1 + 0 \cr
}$$
The fact that $\gcd(20,17) = 1$ tells us that $17$ indeed has an inverse. Now we go backwards to
express $1$ as an integer linear combination of $20$ and $17$:
$$\eqalign{
1 &= 3-1\cdot 2 \cr
&= 3-(17-5\cdot 3) \cr
&= -17 + 6\cdot 3 \cr
&= -17 + 6(20-17) \cr
&= 6\cdot 20 - 7\cdot 17 \cr
}$$
So the inverse of $17$ modulo $20$ is $-7 \equiv_{20} 13$. (We can verify that
$$17\cdot 13\equiv (-3)(-7)\equiv 21 \equiv 1\pmod {20}.$$
As a trick for computations, use the range $-n/2$ to $n/2$ instead of $0$ to $n-1$ to keep the
numbers a bit smaller. It might help you avoid errors.)

Let's find the full list of invertible integers modulo $20$, as well as their inverses. First,
we list the integers in the range $[0,19]$ that are relatively prime with $20$:
$$1,3,7,9,11,13,17,19$$
If some number is in this list, then its inverse must also be in this list (since inverses to
a given element must themselves be invertible).
The elements $1$ and $19$ are their own inverses, since $1\cdot 1 = 1$ and
$19\cdot 19 \equiv_{20} (-1)(-1) = 1$.
We already saw above that $13$ and $17$ are inverses to each other. That leaves $3$, $7$, $9$, and $11$.
First we see that $3\cdot 7 = 21 \equiv_{20} 1$, so $3$ and $7$ are inverses to each other.
Then we note that $9\cdot 9 = 81 \equiv_{20} 1$, so $9$ is its own inverse, leaving us to conclude
that $11$ must also be its own inverse as well---either by the fact that we've ruled
out all other possibilities, or by the computation
$$11\cdot 11 \equiv_{20} (-9)(-9)  = 81 \equiv_{20} 1.$$

\noindent\datestamp{14}{} \hskip\parindent
We now investigate what happens in the particular case where we the modulus $n$ is prime.
In this situation we have the following proposition.

\edef\propinvertiblemodp{\the\thmcount}
\proclaim Proposition \advthm. Let $p$ be prime. Then
\medskip
\item{i)} every $x\in \ZZ$ with $x\not\equiv 0\pmod p$ is invertible modulo $p$; and
\smallskip
\item{ii)} for all $a,b\in \ZZ$ with $ab\equiv 0\pmod p$ one has
$a\equiv 0\pmod p$ or $b\equiv 0\pmod p$.
\medskip

\proof Since $p$ only has factors $1$ and $p$, $\gcd(x,p)$ must either be $1$ or $p$.
But the condition $x\not\equiv 0\pmod p$ implies that $p$ does not divide $x$.
Hence $\gcd(x,p) = 1$, proving part (i).

For part (ii), assume that $ab\equiv 0\pmod p$, so $p\divides ab$. But $p$ is prime, so
by Theorem~{\thmpdivides}, either $p\divides a$ or $p\divides b$. In the first
case we have $a\equiv 0\pmod p$, and in the second case, $b\equiv 0\pmod p$.\slug

\medskip\boldlabel Solving congruences modulo {\mathbold n}.
Suppose we want to solve for all integers $x$ satisfying $x^2 \equiv x\pmod n$, first for $n=6$,
then for $n=7$.
